tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-10 paper.tex

\subsection{Performance}
[START HERE]
\texttt{tstrait} offers a very efficient way of simulating quantitative traits
based on an ARG. Figure \ref{fig:time} (A) shows how the simulation time scales
with the number of individuals on simulated human genomes in \texttt{stdpopsim}
\citep{adrion2020}. The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.

@@

\begin{figure}[!t]%
\centering
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) Computational time with increasing number of individuals. The
whole genome dataset from \emph{Homo sapiens} with various number of
individuals and chromosome length are simulated by using \texttt{stdpopsim}
\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)
Computational time with increasing number of causal sites on two large
realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure}

>>>>

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Fig~\ref{fig:time}A shows how the simulation time scales
with the number of individuals using simple simulations
of humans using \texttt{stdpopsim}~\citep{adrion2020}.
[HERE]
The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.

@@

\begin{figure}[!t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) Computational time with increasing number of individuals. The
whole genome dataset from \emph{Homo sapiens} with various number of
individuals and chromosome length are simulated by using \texttt{stdpopsim}
\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)
Computational time with increasing number of causal sites on two large
realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure}
